Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-11-23 manuscript.tex
Andy markup
Updates from Overleaf

@@

@@

\begin{comment}
%%%%%% intro notes %%%%%%%
\citep{Sanstead2018DirectDehybridization} Good intro with refs and focus on importance of dynamics "rich dynamics even for short oligos" ... Many of the most important functions performed by DNA are highly dynamic, whether in the the biological roles of replication transcription, and gene regulation, or in the expanding field of DNA-based nanotechnology
\end{comment}

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}. Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition. Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates. The stability of certain intermediates states, such as out-of-register or shifted base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}). Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.

Our understanding of DNA dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}. Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it is unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several high-level computational models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.

In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn.2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs. Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, meta-stable structures, and relative timescales.

@@

\subsection{\label{sec:methods}3spn.2 Model}

The 3 site per nucleotide (3spn.2) coarse grain model was designed to accelerate DNA computation relative to all atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. 3spn.2 has been implemented to study DNA pack in viral capsids, protein-DNA binding, and nucleosome unwrapping, and recently served as a basis for the coarser 1cpn model \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping, Lequieu20191CPN:Chromatin}. The model is amenable to an implicit or explicit ion environment -- we chose the latter to include Mg2+ effects from experiment. We achieved better sampling of hybridization events with explicit ions, despite a more demanding computational cost.

@@

We initialized four sequences previously investigated by Sanstead et al. -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We used a Debye-Huckel screening potential to account for phosphate-backbone interactions and preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. We set the simulation about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states. We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time. Base pair coordinates were saved every 30 ps, and full frames including backbones and ions were saved every 3 ns for visualization purposes. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence -- meaning that both the sense and anti-sense strands are identical to each other -- we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system (equation below), providing a robust and object means to evaluate various parameters and models. In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.

@@

% for more general case with  \chi_0, \chi_1, but in practice this is not used
\begin{comment} 
\begin{align}
 	\mathscr{C}_{00}=&\Expect{\chi_0(x_t)\chi_0(x_t)^\intercal}_t\\
 	\mathscr{C}_{01}=&\Expect{\chi_0(x_t)\chi_1(x_{t+\tau})^\intercal}_t\\
 	\mathscr{C}_{11}=&\Expect{\chi_1(x_{t+\tau})\chi_1(x_{t+\tau})^\intercal}_{t+\tau}\\
	\\
 	R_2[\chi_0, \chi_1]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2\\
 	R_2^{val}[\chi_0, \chi_1]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1\\
\end{align}\label{CK2}
\end{comment}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=80mm,
        scale=0.5]{Figs/figs_0804/all_permute_vamps_grouped.png}
        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set. The inverse distances showed improvement across sequences, and the 55 reduced coordinate set performed about as well as larger features set.}
        \label{fig:allseq_features_vamp2}
	\end{center}
\end{figure}

We compared scores for the permutation-free distances, intermolecular, and intra+intermolecular distances and found small differences between the three that varied by sequences (show all sequences in supplemental), but overall we did not observe a loss in kinetic information when using the reduced feature set (\ref{fig:allseq_features_vamp2}). We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal permutation-free coordinates to train the model. The smaller feature set enabled faster training times and better statistics over permutatable distances, without appearing to suffer a loss in generality or model resolution. These features were normalized and passed into sequence-specific SRVs.

% include covariance, vamp-2  and validation equations, lag time equation from vampnet paper mardt2018

\subsubsection{\label{sec:methods}SRVs}

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Molgedey1994SeparationCorrelations, Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

>>>>

%include hyperlinks to figures
\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=false,
    pdfborder={0 0 0},
    linkcolor=blue
]{hyperref}

@@

Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood. In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement. We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Models (MSMs) construction. Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides. For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states. In contrast, G:C pairs near the center of the duplex induce more rapid fraying dynamics. In both cases, hybridization/dissociation mechanisms deviate from an “all-or-nothing” model. Our computational predictions are in excellent accord with experiment, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.

% revise "excellent accord" above?

@@

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. DNA dynamical behavior has become increasingly important in these technologies and a point of focus in fundamental biological processes such as transcription and gene regulation.  \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}. Our understanding of DNA hybridization -- the formation of a duplex from two amenable single strands -- has been integral to these developments. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for many mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}. Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}. Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags altering strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates. For example, metastable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, , Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}). The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequences dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}. Together, these dynamics represent the most substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs. Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

%%%%%%%% Intro comments %%%%%%%%%%%%%%%%%

%Progress in DNA science and engineering has been bolstered by a well-established understanding of sequence-dependent thermodynamics. Thousands of DNA and DNA-complex crystal structures have been captured, and the equilibrium structure of every tetranucleotide motif were simulated in a distributed computational effort \citep{Manuscript2014The2012}. Hybridization thermodynamics can predicted with high fidelity by nearest-neighbor stacking, and these calculations have been experimentally verified to account for mismatched pairs, dangling ends, and other non-native motifs \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}.

% Secondary DNA structures such as hairpins and quadruplexes have also been studied in depth and leverage for nanotechnology applications.

%Furthermore several computational models have been paramaterized to capture these thermodynamic characteristics.

%These studies have fueled increasingly complex structural DNA technologies, but developments in dynamic DNA nanotechnologies have outpaced our understanding of the dynamics themselves \citep{Deluca2020DynamicDevices}. Certain dynamical models for hybridization -- such the nucleation-zipper mechanism and fraying-peeling mechanism for dissociation -- have been well-established for decades (\citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition}). Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

%Still, rich dynamics have been observed short nucleotides \citep{Sanstead2018DirectDehybridization} and the sequence-dependent effects on these dynamics are not well understood.

%Progress has also been made in modeling kinetics for certain dynamic DNA systems, such as toehold exchanges and optical barcodes \citep{Zhang2009ControlExchange, Shah2019ProgrammingFingerprinting}. Sequence effects on association and dissociation rates were predicted using combined experimental and machine learning approaches, however higher order dynamics are less well understood \citep{Zhang2018PredictingSequence}.  Hairpin \citep{Woodside2006DirectAcid} and in duplex hybridization \citep{yer2014KineticsAT-tracts, Sanstead2016, Sanstead2018DirectDehybridization} studies have revealed sequence-dependent effects and varied differential deviations from "all or nothing" behavior. In temperature-jump spectroscopy, G:C placement alters the slow kinetic response attributed to dissociation and a fast response likely corresponding to terminal A:T fraying. Experimentally, however, it is difficult to pick up on dynamics that have similar spectroscopic signatures and to tease apart temporally close responses that might be combined into one stretched exponential signal.

% include Tokmakoff conclusions in the intro -- sequence-dependent deviations from two-state behavior. The slow response was attributed to complete strand dissociation, while the fast response described base fraying. We perform longer equilibrium sampling near Tm to get a sense of sequence equilibrium behavior, as well shorter simulations at a range of higher temperatures to replicate T-jump data. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

@@

The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. 3spn2 has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a A Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states. We use a 20 fs time step and $13\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors for each simulations. Half of runs were initialized from the hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. Depending on sequence, we observed between 40-100 hybridizing and dissociating transitions across all runs.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=80mm,
        scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set. The inverse distances showed improvement across sequences, and the 55-dimension symmetrized coordinate set performed about as well as larger features set.}
        \label{fig:allseq_features_vamp2}
	\end{center}
\end{figure}

We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances and found small differences between the three that varied by sequences, but overall we did not observe a loss in kinetic information when using the symmetrized feature set (\ref{fig:allseq_features_vamp2}). We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal symmetrized coordinates to train the model. The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution. These features were normalized and passed into sequence-specific SRVs.

\subsection{\label{sec:methods}SRVs}

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis},and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

Using optimized hyperparameters and featurized trajectory data, we transformed 55 reciprocal pairwise distances into a low dimensional SRV basis set. In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes. We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}. In particular, we looked for convergence in the VAMP-2 score and inconsistency between cross validation scores -- suggesting that the model may be over-fitting on artifacts in the training data. We used a batch size of 50000 and ran each model for a total of 20 training epochs. We used two hidden layers and set the size of each layer to 100. For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training. SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs. SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.

% might want to include these training params in the SI?

\subsection{\label{sec:methods}SRV-MSMs}

MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results. Additional step-by-step details on SRV-MSM construction for each sequence are provided in the supplemental information and on Github (https://github.com/mrjoness/...).

%%%%%%%% Methods comments %%%%%%%%%%%%%%%%%

%We found close agreement with explicit ion simulations performed with 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} (supplemental) but 7x slower simulation speed made adequate sampling more difficult to achieve under these conditions.

%We initialized explicit ions such that 240 mM \ce{NaCl} and 18 mM \ce{MgCl2} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. We used a periodic box size of 7.774 nm and an effective oligo concentration of 7 mM. We used an Ewald summation to calculate long range Coulombic interaction between DNA and ions using a real space cutoff of 2 nm.

% A Debye-Huckle screening potential was used to monitor account for ion concentration .

\begin{comment} 
% SI equations
In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.

@@
